---
title: "Lilace Scoring Demo"
format:
  gfm:
    toc: true
execute:
  echo: true
  warning: false
  message: false
  eval: true
---

# Use sortscore Output as Lilace Input

The code in this notebook is adapted from the Lilace vignette to demonstrate how to utilize sortscore output as an input for this workflow.

Cite: Freudenberg, J., Rao, J., Howard, M. K., Macdonald, C., Greenwald, N. F., Coyote-Maestas, W., & Pimentel, H. (2026). Accurate variant effect estimation in FACS-based deep mutational scanning data with Lilace. Genome Biology, 27(1), 48. https://doi.org/10.1186/s13059-026-03934-1

Reference vignette: [Introduction to Lilace](https://pimentellab.com/lilace/articles/intro.html).

0. The `sortscore integrate` command is used to return a Lilace-ready counts table.
1. Import scores and metadata from sortscore into Lilace's input format.
2. Normalize according to cell sorting percentages, if needed.
3. Run Lilace.
4. Review the output scores.


## Parameters

Edit these values before running the analysis for a new dataset.

```{r}
#| label: parameters
#| echo: true

# Paths below are repo-relative unless you provide an absolute path.
counts_path <- "demo_data/external_tools/Lilace_input_counts.csv"
working_dir <- "demo_data/external_tools/Lilace"
input_scores_path <- "demo_data/_test_outputs/multitile_output/normalized/zscore_2pole/scores/batch_aa_scores.csv"

# Optional filter for tiled experiments. Set to NULL to include all batches.
batch_name <- "tile1"

# Export type for `sortscore integrate lilace`.
mutagenesis_type <- "aa" # set to "codon" for dna score tables

sortscore_python <- "venv/bin/python" # path to an interpreter with sortscore installed

# Set to TRUE if you want Lilace counts normalized to sorting proportions.
use_sort_normalization <- FALSE

# Sorting proportions aligned to the integrated Lilace count columns (`c_0`, `c_1`, ...).
sort_prop_values <- c(0.25, 0.25, 0.25, 0.25)

control_label <- "synonymous"
n_parallel_chains <- 4
seed <- 1
```

## Installation

The Lilace vignette recommends installing `cmdstanr` first, then `CmdStan`, and finally Lilace itself.

```{r}
#| include: false
if (!requireNamespace("cmdstanr", quietly = TRUE)) {
  install.packages(
    "cmdstanr",
    repos = c("https://stan-dev.r-universe.dev", getOption("repos"))
  )
}
```

```{r}
#| include: false
library(cmdstanr)

if (is.null(cmdstan_version(error_on_NA = FALSE))) {
  install_cmdstan(cores = 2)
}

cmdstan_version()
```

```{r}
#| include: false
if (!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}

if (!requireNamespace("lilace", quietly = TRUE)) {
  remotes::install_github("pimentellab/lilace")
}
```


## Setup

Once the remaining required packages are installed, load them in the setup chunk below.

```{r}
#| include: false
library(here)
here::i_am("demo_data/external_tools/Lilace/Lilace.qmd")
library(readr)
library(dplyr)
library(tidyr)
library(stringr)
library(tibble)
library(lilace)
```

## Generate Input File

### Use sortscore output to get Lilace input counts

Generates the Lilace input counts file from a sortscore scores table.

```{r}
#| message: true
integrate_args <- c(
  "-m", "sortscore", "integrate", "lilace",
  "--input", here::here(input_scores_path),
  "--output", here::here(counts_path),
  "--mutagenesis-type", mutagenesis_type
)

if (!is.null(batch_name) && nzchar(batch_name)) {
  integrate_args <- c(integrate_args, "--batch", batch_name)
}

sortscore_python_path <- if (grepl("^(/|~)", sortscore_python)) {
  path.expand(sortscore_python)
} else {
  here::here(sortscore_python)
}

system2(sortscore_python_path, integrate_args)
```

### Load input counts

The `sortscore integrate lilace` preprocessing step produces one row per variant per replicate, including `variant_id`, `mutation_type`, `position`, and Lilace count columns corresponding to the FACS bins.

```{r}
counts_file <- here::here(counts_path)
working_dir_path <- here::here(working_dir)

raw_counts <- readr::read_csv(counts_file, show_col_types = FALSE)

dplyr::glimpse(raw_counts)
head(raw_counts)
```


### Prepare counts for Lilace

The integrated input file already includes Lilace's `variant_id`, `mutation_type`, `position`, `replicate`, and `c_0`, `c_1`, ... count columns. The step below detects those count columns and separates any extra metadata.

```{r}
stopifnot(all(c("variant_id", "mutation_type", "position", "replicate") %in% names(raw_counts)))

count_col_names <- names(raw_counts)[grepl("^c_[0-9]+$", names(raw_counts))]
stopifnot(length(count_col_names) > 0)

lilace_counts <- raw_counts |>
  dplyr::select(
    variant_id,
    mutation_type,
    position,
    replicate,
    dplyr::all_of(count_col_names)
  )

lilace_counts[count_col_names] <- lapply(
  lilace_counts[count_col_names],
  function(col) as.integer(round(col))
)

metadata_cols <- raw_counts |>
  dplyr::select(-replicate, -dplyr::all_of(count_col_names))

lilace_counts |>
  dplyr::arrange(position, variant_id, replicate) |>
  head()

metadata_cols |>
  dplyr::arrange(position, variant_id) |>
  head()
```

Convert the counts table into a Lilace object.

```{r}
lilace_obj <- lilace::lilace_from_counts(
  variant_id = lilace_counts$variant_id,
  mutation_type = lilace_counts$mutation_type,
  position = lilace_counts$position,
  replicate = lilace_counts$replicate,
  counts = lilace_counts |> dplyr::select(dplyr::all_of(count_col_names)),
  metadata = metadata_cols
)

head(lilace_obj$data)
```

## Run Lilace

Optionally normalize to the FACS sorting proportions.

```{r}
#| label: normalize sorting proportions
if (use_sort_normalization) {
  stopifnot(length(sort_prop_values) == length(count_col_names))
  sort_props <- stats::setNames(sort_prop_values, count_col_names)

  lilace_obj <- lilace::lilace_sorting_normalize(
    lilace_obj,
    sort_props,
    rep_specific = FALSE
  )
}

if (use_sort_normalization) {
  head(lilace_obj$normalized_data)
}
```

Fit the Lilace model. Per the Lilace vignette, synonymous variants are used as the negative control group by default.

```{r}
#| label: fit model
#| warning: true
#| message: true
lilace_obj <- lilace::lilace_fit_model(
  lilace_obj,
  output_dir = working_dir_path,
  control_label = control_label,
  control_correction = TRUE,
  use_positions = TRUE,
  pseudocount = TRUE,
  n_parallel_chains = n_parallel_chains,
  seed = seed
)
```

Read the exported scores table.

```{r}
#| label: read scores
scores <- readr::read_tsv(
  file.path(working_dir_path, "lilace_output", "variant_scores.tsv"),
  show_col_types = FALSE
)

head(scores)
```

Basic summary views.

```{r}
#| label: summarize
scores |>
  dplyr::count(type, sort = TRUE)

scores |>
  dplyr::arrange(lfsr, dplyr::desc(abs(effect))) %>%
  dplyr::select(variant, type, position, effect, effect_se, lfsr, discovery05) %>%
  head(20)
```

Optional diagnostic plots from Lilace.

```{r}
#| label: effect size heatmap
lilace::lilace_score_density(
  scores,
  savedir = working_dir_path,
  score.col = "effect",
  name = "score_histogram",
  hist = TRUE,
  scale.free = TRUE,
  savepdf = FALSE
)

heatmap_scores <- scores |>
  dplyr::mutate(
    wildtype = stringr::str_extract(variant, "^[^.]+"),
    mutation = stringr::str_extract(variant, "[^.]+$")
  )

lilace::lilace_score_heatmap(
  heatmap_scores,
  savedir = working_dir_path,
  score.col = "effect",
  name = "score_heatmap",
  x.text = 9,
  y.text = 10,
  seq.text = 3,
  savepdf = FALSE
)
```

```{r}
#| label: discovery heatmap
lilace::lilace_score_heatmap(
  heatmap_scores,
  savedir = working_dir_path,
  score.col = "discovery05",
  name = "discovery05_heatmap",
  x.text = 9,
  y.text = 10,
  seq.text = 3,
  savepdf = FALSE
)
```
